---
title: "Fitbit Daily Activity: Behaviroal Patterns and Energy Expenditure"
author: "Mitchell Talton"
output: 
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true 
---
## Overview
This notebook explores daily activity patterns from a public Fitbit dataset, focusing on:
1) **Behavioral variability** (weekday vs weekend step patterns)  
2) **Physiological relationship** (steps vs calories burned)

To improve interpretability, analysis is performed at the **individual-user** level.

## Load Libraries

```{r}
library(tidyverse)
library(lubridate)

theme_set(theme_minimal(base_size=13))
```

## Load Data

```{r}
daily_data <- read_csv("dailyActivity_merged.csv", show_col_types = FALSE)

glimpse(daily_data)

```
## Data Cleaning Data and Feature Engineering
```{r}
daily_data <- daily_data %>%
  mutate(
    ActivityDate = as.Date(ActivityDate, format = "%m/%d/%Y"),
    DayOfWeek = wday(ActivityDate, label = TRUE, abbr = FALSE),
    DayType = if_else(DayOfWeek %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
  )
```

```{r}
stopifnot(!all(is.na(daily_data$ActivityDate)))
summary(daily_data$TotalSteps)
```

## Single-User Selection

```{r}
#Wearable datasets often contain multiple participants. For clear visual storytelling, I will focus on one user.
user_id <- daily_data$Id %>% unique() %>% .[1]

user_data <- daily_data %>%
  filter(Id == user_id) %>%
  filter(!is.na(TotalSteps), !is.na(Calories)) %>%
  arrange(ActivityDate)

range(user_data$ActivityDate)

```

## Behavioral Pattern: Steps by Day Type
**Question:** Do step counts differ between weekdays and weekends for this user? 
```{r}
ggplot(user_data, aes(x = ActivityDate, y = TotalSteps, color = DayType)) +
  geom_point(alpha = 0.65) +
  geom_smooth(se = FALSE, method = "loess") +
  labs(
    title = "Daily Step Count by Day Type",
    subtitle = paste("User:", user_id),
    x = "Date",
    y = "Total Steps",
    color = "Day Type"
  )
```
**Interpretation:** Weekday activity is more consistent, while weekend activity shows greater variability—suggesting differences in routine structure across the week.

## Physiological Context: Steps vs Calories
**Question:** How strongly does locomotor activity (steps) relate to energy expenditure (calories)?

```{r}
user_eff <- user_data %>%
  filter(TotalSteps >= 2000) %>%  
  mutate(CaloriesPerStep = Calories / TotalSteps)

summary(user_eff$CaloriesPerStep)

ggplot(user_eff, aes(x = DayType, y = CaloriesPerStep)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.5) +
  geom_jitter(width = 0.12, alpha = 0.6) +
  labs(
    title = "Energy Cost per Step by Day Type",
    subtitle = "Calories/step (proxy for intensity or non-step expenditure); low-step days removed",
    x = "Day Type",
    y = "Calories per Step"
  ) +
  theme_minimal()

user_eff %>%
  group_by(DayType) %>%
  summarise(
    Days = n(),
    Mean_CalPerStep = mean(CaloriesPerStep),
    Median_CalPerStep = median(CaloriesPerStep),
    .groups = "drop"
  )
```
**Interpretation:**Calories per step serves as a simple proxy for energy cost per unit movement. Differences by day type may reflect changes in activity intensity (e.g., more vigorous movement on weekends) or additional non-step energy expenditure. Low-step days were excluded to reduce ratio inflation.

## Quantifying Weekday vs. Weekend Differences

```{r}
library(scales)

summary_table <- user_eff %>%
  group_by(DayType) %>%
  summarise(
    Days = n(),
    MeanSteps = mean(TotalSteps),
    MedianSteps = median(TotalSteps),
    MeanCalories = mean(Calories),
    MedianCalories = median(Calories),
    MeanCalPerStep = mean(CaloriesPerStep),
    MedianCalPerStep = median(CaloriesPerStep),
    .groups = "drop"
  )

summary_table

```

```{r}
# Wide format for quick comparisons
diff_table <- summary_table %>%
  select(DayType, MeanSteps, MeanCalories, MeanCalPerStep) %>%
  pivot_wider(names_from = DayType, values_from = c(MeanSteps, MeanCalories, MeanCalPerStep)) %>%
  mutate(
    Steps_Diff = MeanSteps_Weekend - MeanSteps_Weekday,
    Calories_Diff = MeanCalories_Weekend - MeanCalories_Weekday,
    CalPerStep_Diff = MeanCalPerStep_Weekend - MeanCalPerStep_Weekday
  )

diff_table


```
**Takeaway:** Weekday vs weekend differences are summarized above to support interpretation of behavioral structure (steps) and relative energy cost (calories/step).

## Activity Consistency 
```{r}
consistency <- user_data %>%
  group_by(DayType) %>%
  summarise(
    Days = n(),
    MeanSteps = mean(TotalSteps, na.rm = TRUE),
    SDSteps = sd(TotalSteps, na.rm = TRUE),
    CVSteps = SDSteps / MeanSteps,
    .groups = "drop"
  )

consistency
```
**Interpretation:** CV (coefficient of variation) reflects consistency. Lower CV indicates more stable routine-driven activity patterns.

```{r}
ggplot(user_data, aes(x = DayType, y = TotalSteps)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.5) +
  geom_jitter(width = 0.12, alpha = 0.6) +
  labs(
    title = "Step Count Distribution by Day Type",
    subtitle = paste("User:", user_id),
    x = "Day Type",
    y = "Total Steps"
  )

```

## Calories per Step vs. Steps:
```{r}
ggplot(user_eff, aes(x = TotalSteps, y = CaloriesPerStep, color = DayType)) +
  geom_point(alpha = 0.7) +
  geom_smooth(se = FALSE, method = "loess") +
  labs(
  title = "Calories per Step Decreases with Higher Daily Step Counts",
  subtitle = "Reflects baseline energy expenditure being distributed across greater movement volume",
  x = "Total Steps",
  y = "Calories per Step",
  color = "Day Type"
)
```
**Interpretation:**Calories per step decreases as daily step count increases because total daily energy expenditure includes baseline metabolic processes (e.g., BMR and non-locomotor activity). As step count rises, this fixed baseline is distributed across greater movement volume, lowering the calories-per-step ratio. This demonstrates a limitation of ratio-based metrics and reinforces the importance of separating behavioral volume (steps) from physiological cost when interpreting wearable data.

## Key Findings (Single-User View)

**Behavioral structure:** Weekday step counts were more consistent, while weekends exhibited greater variability, suggesting differences in routine-driven activity patterns.
**Energy context:** Calories per step differed modestly by day type after excluding low-step days, reflecting how baseline metabolic expenditure interacts with activity volume.
**Methodological insight:** Separating behavioral metrics (steps) from physiological context (energy expenditure) reduced misleading artifacts and improved interpretability of wearable data.